---
title: "hycom"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Provides for access from R to [HYCOM](https://www.hycom.org/) online datasets.

### [Citation](https://www.hycom.org/)

>Bleck, R., (2001). An oceanic general circulation model framed in hybrid isopycnic-Cartesian coordinates. Ocean Modeling, 4: 55-88.

### Requirements

From CRAN...

  + [R v4+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [sf](https://CRAN.R-project.org/package=sf)
  + [stars](https://CRAN.R-project.org/package=stars)

From github...

  + [xyzt](https://github.com/BigelowLab/xyzt)

### Installation

```
remotes::install_github("BigelowLab/hycom")
```

### Getting started

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(sf)
  library(hycom)
  library(xyzt)
  library(stars)
})
```

#### Working with points.  

See the [xyzt](https://github.com/BigelowLab/xyzt) package for more details on the example Southern US Atlantic Bight data.

```{r points}
# read in example SAB points
x <- xyzt::read_sab() |>
  dplyr::mutate(lon = xyzt::to_360(lon),  # HYCOM lons are [0,360]
                depth = abs(depth)) |>    # HYCOM depths are positive downward
  xyzt::as_POINT(dims = 'xyzt')

# generate a hycom url for a dataset
url <- hycom_url(filename = "ts3z")

# open the resource
X <- ncdf4::nc_open(url)

# extract the data 
covars <- hycom::extract(x, X, varname = c("salinity", "water_temp"))


# bind to the input
(y <- dplyr::bind_cols(x, covars))
```

#### Working with bounding boxes (from points or polygons). 

Learn more about working with [stars](https://CRAN.R-project.org/package=stars) objects in the [vignettes](https://r-spatial.github.io/stars/).

```{r boxes}
# read in example SAB points
x <- xyzt::read_sab() |>
  dplyr::mutate(lon = xyzt::to_360(lon),
                depth = abs(depth)) |>    
  xyzt::as_POINT(dims = 'xyzt') |>
  xyzt::as_BBOX()

(covars <- hycom::extract(x, X, varname = c("salinity", 'water_temp')))
```

Now let's see what it looks like.

```{r show_boxes}
x <- xyzt::read_sab() |>
  dplyr::select(-time, -depth) |>
  dplyr::mutate(lon = xyzt::to_360(lon)) |>
  xyzt::as_POINT()


par(mfrow = c(1,2))
plot(covars['salinity'], key.pos = NULL, axes = TRUE, reset = FALSE)
plot(sf::st_geometry(x), add = TRUE, col = "orange", pch = 19, cex = 2)

plot(covars['water_temp'], key.pos = NULL, axes = TRUE, reset = FALSE)
plot(sf::st_geometry(x), add = TRUE, col = "orange", pch = 19, cex = 2)

```



```{r cleanup}
# cleanup
ncdf4::nc_close(X)
```
