---
title: "hycom"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Provides for access from R to [HYCOM](https://www.hycom.org/) online datasets.

### [Citation](https://www.hycom.org/)

>Bleck, R., (2001). An oceanic general circulation model framed in hybrid isopycnic-Cartesian coordinates. Ocean Modeling, 4: 55-88.

### Requirements

From CRAN...

  + [R v4+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [sf](https://CRAN.R-project.org/package=sf)
  + [stars](https://CRAN.R-project.org/package=stars)

### Installation

```
remotes::install_github("BigelowLab/hycom")
```

### Getting started

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(sf)
  library(stars)
  library(hycom)
})
```

### Working with points and bounding boxes  

See the [xyzt](https://github.com/BigelowLab/xyzt) package for more details on the example US South Atlantic Bight (SAB) data; the location data following is adapted from that package.  Here we make a bounding box and a selection of points.  We also load a coastline from [NaturalEarth](https://www.naturalearthdata.com/) to use with plotting.

Note that the data frame, `bouys`, must be converted to a sf object, which we call `points` to work with other sf objects. HYCOM is a 4 dimensional dataset of `lon`, `lat`, `depth` and `time` - sometimes this is called XYZT or XYZM (M for some other dimension other than time).  sf objects handle all of the details for us, but we must inform sf of the names of our 4 dimensions.   

```{r points}
coast = rnaturalearth::ne_coastline(returnclass = 'sf', scale = "large")
box = sf::st_bbox( c(xmin = -81, ymin = 27, xmax = -71, ymax = 35), crs = 4326)
buoys = dplyr::tribble(
      ~id,           ~name,    ~lon,   ~lat, ~depth,               ~time,
  "41009",      "Canveral",   -80.2,   28.5,    1.5, "2022-06-17 18:05:06",
  "41010","Canaveral East",   -78.5,   28.9,    1.5, "2022-06-29 18:05:06",
  "41002","South Hatteras",   -74.9,   31.8,    1.5, "2022-06-05 18:05:06",
  "41001", "East Hatteras",   -72.2,   34.7,    2.0, "2022-04-30 18:05:06",
  "41004",        "EDISTO",   -79.1,   32.5,    1.5, "2022-07-11 18:05:06") |>
  dplyr::mutate(time = as.POSIXct(time, format = "%Y-%m-%d %H:%M:%S", tz = 'UTC'))

points = buoys |>
  sf::st_as_sf(coords = c("lon", "lat", "depth", "time"), dim = "XYZM",crs = 4326)
points
```

Note that the `geometry` column contains 4 elements (XYZM) but also note that time has reverted to the number of seconds since the time-origin of 1970-01-01T00::00:00Z.

Now we can plot these with the coast.

```{r}
plot(points['name'], axes = TRUE, main = "South Atlantic Bight", 
     extent = box, reset = FALSE, key.width = lcm(8), pch = 16, cex = 2)
plot(sf::st_geometry(coast), add = TRUE, border = "black")
```

### Accessing data


#### Generate a hycom url for a dataset

You have to determine which file you want to access; currently we can access this [suite of products](https://tds.hycom.org/thredds/catalogs/GLBy0.08/expt_93.0.html). The `hycom_url()` function will craft the URL once you decide what file you want to access.

```{r}
url = hycom_url(filename = "ts3z",
                product = "GLBy0.08",
                version = "expt_93.0")
url
```



#### Open the resource

And then open the resource - it's a NetCDF.  Printed below is a summary of the contents of the resource including the variables and the dimensions over which the are defined.
```{r}
X <- ncdf4::nc_open(url)
X
```


#### An important point

Data provided by HYCOM are mapped to range of longitude [0, 360], but most oceanography work is done in the range [-180, 180].  To extract data, points or rasters, we must first transform our longitudes to the [0,360] range.  We provide functions, `to_180()` and `to_360()` to facilitate this transformation.  Below we transform make new datasets form our previous datasets.

```{r}
buoys_360 = buoys |>
  dplyr::mutate( lon = to_360(lon))
points_360 = buoys_360 |>
  sf::st_as_sf(coords = c("lon", "lat", "depth", "time"), dim = "XYZM",crs = 4326)
box_360 = sf::st_bbox( c(xmin = to_360(-81), 
                         ymin = 27, 
                         xmax = to_360(-71), 
                         ymax = 35), 
                       crs = 4326)
```

#### Extract the data 

Here we extract salinity and water_temp and bind these to the points.
```{r}
covars <- hycom::extract(points_360, X, varname = c("salinity", "water_temp"))
# bind to the input
(y <- dplyr::bind_cols(points, covars))
```

#### Working with bounding boxes (from points or polygons). 

Learn more about working with [stars](https://CRAN.R-project.org/package=stars) objects in the [vignettes](https://r-spatial.github.io/stars/). First need to select a time and depth.

```{r time}
time = hycom_time(X)
cat("head of time\n")
head(time)
cat("tail of time\n")
tail(time)
```
It looks like every three hours since Dec 4, 2018.  We can just pick time in that range (it doesn't have to be extact as we interpolate.)

```{r depth}
depth = hycom_depth(X)
cat("head of depth\n")
head(depth)
cat("tail of depth\n")
tail(depth)
```
That is irregular spacing, but once again we can request any depth and the software will interpolate to one of the set depths.

```{r boxes}
covars <- extract(box_360, X, 
                  time = as.POSIXct("2020-01-06 09:00:00", tz = 'UTC'),
                  depth = 2000,
                  varname = c('salinity', 'water_temp'))
covars
```

Now let's see what it looks like.

```{r show_boxes}
par(mfrow = c(1,2))
plot(covars['salinity'], key.pos = NULL, axes = TRUE, reset = FALSE)
plot(sf::st_geometry(points_360), add = TRUE, col = "orange", pch = 19, cex = 2)

plot(covars['water_temp'], key.pos = NULL, axes = TRUE, reset = FALSE)
plot(sf::st_geometry(points_360), add = TRUE, col = "orange", pch = 19, cex = 2)

```



```{r cleanup}
# cleanup
ncdf4::nc_close(X)
```
